---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = ">",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# ox

The goal of the `ox` package is to read OpenClinica odm 1.3 xml export files, 
and create dataframes for data and metadata (or an `ox_all` object, a list 
containing both) from parsed xml files. Also, tidy dataframes can be created for any ItemGroup, with optional definition of factors from codelists.

`ox` is currently under development.

## Installation

You can install the development version of `ox` from [GitHub](https://github.com/acobos/ox) with:

``` r
install_github("acobos/ox")
```

## Example

This is a basic example which shows you how to solve a common problem: create a 
tidy dataframe for a group of related items (an **ItemGroup** in OpenClinica 
terminology), e.g., for demographic data.

First, load packages `XML`and `ox`. Then parse the xml file using `xmlParse` 
from package `XML`, and use function `ox_all()` to create an `ox_all` object. 
This involves reading data and metadata, and is a slow process. It may take a 
lot of time with any real study!

With `ox_info()` you will see some basic information of this object.

```{r example}

# load packages 
library(XML)
library(ox)

# the xml file address/name
xml_file <- system.file("extdata",
                        "odm1.3_full_example.xml",
                        package = "ox",
                        mustWork = TRUE)

# parse the xml file
doc <- xmlParse(xml_file) 

# create an (ox_all) object 
d <- ox_all(doc)

# get basic info
ox_info(d)

```

In the previous output, `$numbers` documents that this example data includes 240 datapoints on two subjects, from two sites, etc.  

One of the ItemGroups in this study (`$groups` in the output above) is 
`IG_DEMO_DEMOGRAPHICDATA`. Let's get a tidy dataframe for this group, and see 
the result. 

```{r example_cont}
# get tidy dataframe for demographic data
demo <- ox_xtract_group(d, group = "IG_DEMO_DEMOGRAPHICDATA",
                        use_item_names = TRUE,
                        define_factors = TRUE)

demo
```

In the resulting dataframe, columns `study_oid` to `group_repeat_key` are keys for study site, subject, event, etc.; `demo_age` and `demo_menstrual` are the relevant variables, i.e., the two items included in the `IG_DEMO_DEMOGRAPHICDATA` ItemGroup. 

See the vignette for more details. 
